Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-05-07 ppdp19/ppdp19.tex
[ ppdp19 ] Crude 2-column formatting done

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen]{acmart}

@@

@@

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
\citestyle{acmauthoryear}

@@

@@

We can define
\fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
\sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
by recursion, meaning that the (proof-relevant)
sublist relation is reflexive and transitive.
% The sublist relation is reflexive and transitive, the witnesses
% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
% \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
% have easy recursive definitions.
Thus, lists $\Gamma$ form a category
$\Cxt$ with morphisms $\tau : \Gamma \sublist \Delta$, and the category laws
hold propositionally, e.g., we have $\id \rcomp \tau \equiv \tau$ in
propositional equality for all morphisms $\tau$.
The singleton weakening $\twk^A_\Gamma : \Gamma \sublist \cext \Gamma A$,
also written $\twk^A$ or $\twk$, is defined by $\twk = \tweak\,\tid$.

@@

Terms of type $A$ form a presheaf
$A{\from}\_$ as witnessed by the standard weakening operation%
\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
calculus we should better speak of \emph{reindexing}, which could, a
bit clumsily, be also abbreviated to $\tren$.}
\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
defined by recursion over $t : A \from \Gamma$,
and functor laws for $\tren$ analogously to $\treindex$.

@@

As expected, $\tNe\,A$ and $\tNf\,A$ are presheaves,
\ie, support reindexing with $\tren$ just as terms do.
From a normal form we can extract the term via an overloaded function
\fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
\fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
that discards constructor $\tne$ but keeps all other constructors.
This erasure function naturally commutes with reindexing, making it a
natural transformation between the presheaves $\tNf\,A$ ($\tNe\,A$,
resp.) and $A{\from}\_$.  We shall simply write, for instance,
$\tNf\,A \todot A{\from}\_$ for such presheaf morphisms.
(The point on the arrow is mnemonic for \emph{pointwise}.)
Slightly abusive, we shall extend this notation to $n$-ary
morphisms, \eg, write $\A \todot \B \todot \C$ for
$\forall \Gamma.\ \A\,\Gamma \to (\B\,\Gamma \to \C\,\Gamma)$.

@@

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{l@{~}c@{~}l@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
\\ &
  & \mbox{case on absurd neutral} \\
\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
\\
  &\to& \C\,\B\,(\cext \Gamma {A_1})
  \to \C\,\B\,(\cext \Gamma {A_2})
  \to \C\,\B\,\Gamma
\\ &
  & \mbox{case on neutral} \\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
\\ &
  & \mbox{run the monad ($\tNf$ only)} \\
\end{array}
\]
To make things concrete,
we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined
%inductively by
as an inductive family with constructors
$\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
and $A_2$, resp.  Leaves are either
labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind
(aka Kleisli extension) replaces
these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$
which flattens a 2-level case tree, \ie, a case tree with case trees as
leaves, into a single one.
Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ shall be constructed by the
following rules:
\begin{gather*}
  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
It repeats the introductions and eliminations of negative types,
except that application is restricted to values.  Values of type
$\Thunk N$ are embedded via $\tforce$.
Further, values of type $P$ can embedded
via $\tret$, producing a term of type $\Comp P$.  Such terms are
eliminated by $\tbind$ which is, unlike the usual monadic bind, not
only available for $\tComp$-types but for arbitrary negative types
$N$.  This is justified by the monadic character of negative
types, by virtue of $\trun^N$.  Finally, there are eliminators
($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
sum types.
\begin{gather*}
  \nru{\tret}
      {\PValTm P \Gamma}
      {\TmP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Tm N {(\cext \Gamma P)}}
      {\TmP {P \To N} \Gamma}
\quad
  \nru{\npair}
      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
      {\TmP {N_1 \ntimes N_2} \Gamma}
\quad
  \nru{\nunit}
      {}
      {\Tm \top \Gamma}
\\[2ex]
  \nru{\tforce}
      {\PValTmP {\Thunk N} \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tapp}
      {\TmP {P \To N} \Gamma \qquad \PValTm P \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tprj_i}
      {\TmP {N_1 \ntimes N_2} \Gamma}
      {\Tm {N_i} \Gamma}
\\[2ex]
% \end{gather*}
% \begin{gather*}
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
      {\Tm N \Gamma}
\qquad
  \nru{\tsplit}
    {\PValTmP {P_1 \times P_2} \Gamma
      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
    }{\Tm N \Gamma}
\\[2ex]
  \nru{\tcase}
    {\PValTmP {P_1 + P_2} \Gamma
      \qquad \Tm N {(\cext \Gamma P_1)}
      \qquad \Tm N {(\cext \Gamma P_2)}
    }{\Tm N \Gamma}
\qquad
  \nru{\tabort}
    {\PValTm \tyzero \Gamma}
    {\Tm N \Gamma}
\end{gather*}

Interpretation of values
$\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
and terms
$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
is straightforward, thanks to the pioneering work of
Moggi \cite{moggi:infcomp91} and
Levy \cite{levy:hosc06}
put into the design of CBPV.
\[
\begin{array}{lcl}
  \\
  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
  \dent{\punit}\gamma   & = & \ttempty \\
  \dent{\ppair\,v_1\,v_2}\gamma & = &
    (\dent{v_1}\gamma,\ \dent{v_2}\gamma) \\
  \dent{\inj i v}\gamma & = & \iota_i \dent v \gamma \\
  \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
\end{array}
\quad
\begin{array}{lcl}
  \dent{\tabs\,t}\gamma
      & = & \fdent t \gamma \\
  \dent{\tapp\,t\,v} \gamma & = &
     \dent t \gamma\;\tid\;\dent v \gamma \\
  \dent{\nunit}\gamma & = & \ttempty \\
  \dent{\npair\,t_1\,t_2}\gamma & = &
    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
\end{array}
\]
Since $\tThunk$ serves only as an embedding of negative into positive
types and has no semantic effect,
we interpret thunking and forcing by the identity.
The eliminations for positive types deal now only with values, thus,
need not reference the monad operations.
\[
\begin{array}{lcl}
  \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
    \dent{t}{(\gamma,a_1,a_2)} ) \; \dent v \gamma \\
  \dent{\tcase\,v\,t_1\,t_2}\gamma & = &
    [  \lambda a_1.\ \dent{t_1}{(\gamma,a_1)}
    ,\ \lambda a_2.\ \dent{t_2}{(\gamma,a_2)}
    ]\; \dent v \gamma \\
  \dent{\tabort\,v}\gamma & = & \tmagic\,\dent v \gamma \\
\end{array}
\]
The use of the monad is confined to $\tret$ and $\tbind$.
Note the availability of $\trun^\C : \C \den N \to \den N$
at any negative type $N$ for the interpretation of $\tbind$.
\[
\begin{array}{lcl}
  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
\end{array}
\]

@@

Positive normal forms are values $v : \VNf P \Gamma$
referring only to atomic variables and
whose thunks only
contain negative normal forms.
% Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
% contain negative normal forms.
\begin{gather*}
  \nru{\tvar}{\patom \in \Gamma}{\PValNf {\patom} \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValNf P \Gamma}
\qquad
  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValNf{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValNf{P_1}\Gamma \qquad \PValNf{P_2}\Gamma}
      {\PValNfP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
\end{gather*}

Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
starting from a forced $\tThunk$ rather than from variables of negative
types (as those do not exist in CBPV).  However, due to normality
the $\tThunk$ cannot be a $\tthunk$, but only a variable
$\Thunk N \in \Gamma$.
\begin{gather*}
  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeP {N_1 \ntimes N_2} \Gamma}
      {\Ne {N_i} \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\end{gather*}
Variables are originally introduced by either $\tabs$ or
the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
$P$.  Variables of composite value type can be broken down by pattern
matching, introducing variables of smaller type.  These positive
eliminations plus $\tbind$ are organized in the
inductively defined strong monad \fbox{$\Cov$}.
In the following rules, parameter $\J$ (for ``judgement'') stands for an
arbitrary presheaf.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tsplit}
      {P_1 \times P_2 \in \Gamma \qquad
       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tcase}
      {P_1 + P_2 \in \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\qquad
  \nru{\tabort}
      {\tyempty \in \Gamma}
      {\Cover \J \Gamma}
\end{gather*}
Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
$\Comp P$ is reached.  Then, elimination of neutrals and variables is
possible through the $\Cov$ monad until an answer can be given in form
of a base neutral ($\tNe\,\natom$) or a normal value.
\begin{gather*}
  \nru{\tne}
      {\CoverP {\tNe\,\natom} \Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tret}
      {\CoverP {\tVNf\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\\[2ex]
  \nru{\nunit}{}{\Nf \top \Gamma}
\qquad
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\qquad
  \nru{\tabs}
      {\Nf N {\cext \Gamma P}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
We again can run the cover monad on normal forms, \ie, have
$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
negative semantic values
$\trun : \Cov \den N \todot \den N$.

>>>>

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,screen,fleqn]{acmart}

@@

%%\usepackage{nccmath} %ceqn  %% NOT COMPATIBLE

@@

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
% \citestyle{acmauthoryear}

@@

\setlength{\mathindent}{\parindent}

@@

We can define
\[
\begin{array}{lcl}
\tid & : & \Gamma \sublist \Gamma
\\
\_{\rcomp}\_ & : & \Gamma
\sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi
\end{array}
\]
% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
% \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
by recursion, meaning that the (proof-relevant)
sublist relation is reflexive and transitive.
% The sublist relation is reflexive and transitive, the witnesses
% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
% \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
% have easy recursive definitions.
Thus, lists $\Gamma$ form a category
$\Cxt$ with morphisms $\tau : \Gamma \sublist \Delta$, and the category laws
hold propositionally, e.g., we have $\id \rcomp \tau \equiv \tau$ in
propositional equality for all morphisms $\tau$.
The singleton weakening $\twk^A_\Gamma : \Gamma \sublist \cext \Gamma A$,
also written $\twk^A$ or $\twk$, is defined by $\twk = \tweak\,\tid$.

@@

Terms of type $A$ form a presheaf
$A{\from}\_$ as witnessed by the standard weakening operation%
\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
calculus we should better speak of \emph{reindexing}, which could, a
bit clumsily, be also abbreviated to $\tren$.}
%\fbox{$
\[
\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta
\]
%$}
defined by recursion over $t : A \from \Gamma$,
and functor laws for $\tren$ analogously to $\treindex$.

@@

As expected, $\tNe\,A$ and $\tNf\,A$ are presheaves,
\ie, support reindexing with $\tren$ just as terms do.
From a normal form we can extract the term via an overloaded function
\[
\begin{array}{lcl}
\tm\_ & : & \Nf A \Gamma \to A \from \Gamma \\
\tm\_ & : & \Ne A \Gamma \to A \from \Gamma \\
\end{array}
\]
% \fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
% \fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
that discards constructor $\tne$ but keeps all other constructors.
This erasure function naturally commutes with reindexing, making it a
natural transformation between the presheaves $\tNf\,A$ ($\tNe\,A$,
resp.) and $A{\from}\_$.  We shall simply write, for instance,
$\tNf\,A \todot A{\from}\_$ for such presheaf morphisms.
(The point on the arrow is mnemonic for \emph{pointwise}.)
Slightly abusive, we shall extend this notation to $n$-ary
morphisms, \eg, write $\A \todot \B \todot \C$ for
$\forall \Gamma.\ \A\,\Gamma \to (\B\,\Gamma \to \C\,\Gamma)$.

@@

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{l@{~}c@{~}l@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
%\\ & & \mbox{case on absurd neutral}
\\
\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
\\
  && {} \to \C\,\B\,(\cext \Gamma {A_1})
  \to \C\,\B\,(\cext \Gamma {A_2})
  \to \C\,\B\,\Gamma
%\\ & & \mbox{case on neutral}
\\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
%\\ & & \mbox{run the monad ($\tNf$ only)}
\\
\end{array}
\]
To make things concrete,
we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined
%inductively by
as an inductive family with constructors
$\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
and $A_2$, resp.  Leaves are either
labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind
(aka Kleisli extension) replaces
these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$
which flattens a 2-level case tree, \ie, a case tree with case trees as
leaves, into a single one.
Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ shall be constructed by the
following rules:
\begin{gather*}
  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\\[2ex]
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
It repeats the introductions and eliminations of negative types,
except that application is restricted to values.  Values of type
$\Thunk N$ are embedded via $\tforce$.
Further, values of type $P$ can embedded
via $\tret$, producing a term of type $\Comp P$.  Such terms are
eliminated by $\tbind$ which is, unlike the usual monadic bind, not
only available for $\tComp$-types but for arbitrary negative types
$N$.  This is justified by the monadic character of negative
types, by virtue of $\trun^N$.  Finally, there are eliminators
($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
sum types.
\begin{figure}[htbp]
%\begin{ceqn}
\hrulefill
\begin{gather*}
  \nru{\tret}
      {\PValTm P \Gamma}
      {\TmP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Tm N {(\cext \Gamma P)}}
      {\TmP {P \To N} \Gamma}
\\[2ex]
  \nru{\nunit}
      {}
      {\Tm \top \Gamma}
\quad
  \nru{\npair}
      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
      {\TmP {N_1 \ntimes N_2} \Gamma}
\\[2ex]
  \nru{\tforce}
      {\PValTmP {\Thunk N} \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tapp}
      {\TmP {P \To N} \Gamma \qquad \PValTm P \Gamma}
      {\Tm N \Gamma}
\\[2ex]
  \nru{\tprj_i}
      {\TmP {N_1 \ntimes N_2} \Gamma}
      {\Tm {N_i} \Gamma}
\\[2ex]
% \end{gather*}
% \begin{gather*}
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
      {\Tm N \Gamma}
\\[2ex]
  \nru{\tsplit}
    {\PValTmP {P_1 \times P_2} \Gamma
      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
    }{\Tm N \Gamma}
\\[2ex]
  \nru{\tcase}
    {\PValTmP {P_1 + P_2} \Gamma
      \qquad \Tm N {(\cext \Gamma P_1)}
      \qquad \Tm N {(\cext \Gamma P_2)}
    }{\Tm N \Gamma}
\\[2ex]
  \nru{\tabort}
    {\PValTm \tyzero \Gamma}
    {\Tm N \Gamma}
\end{gather*}
%\end{ceqn}
\hrulefill
  \caption{Terms of CBPV}
  \label{fig:tmcbpv}
\end{figure}

Interpretation of values
$\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
and terms
$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
is straightforward, thanks to the pioneering work of
Moggi \cite{moggi:infcomp91} and
Levy \cite{levy:hosc06}
put into the design of CBPV.
\[
\begin{array}{lcl}
  \\
  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
  \dent{\punit}\gamma   & = & \ttempty \\
  \dent{\ppair\,v_1\,v_2}\gamma & = &
    (\dent{v_1}\gamma,\ \dent{v_2}\gamma) \\
  \dent{\inj i v}\gamma & = & \iota_i \dent v \gamma \\
  \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
% \end{array}
% \quad
% \begin{array}{lcl}
  \dent{\tabs\,t}\gamma
      & = & \fdent t \gamma \\
  \dent{\tapp\,t\,v} \gamma & = &
     \dent t \gamma\;\tid\;\dent v \gamma \\
  \dent{\nunit}\gamma & = & \ttempty \\
  \dent{\npair\,t_1\,t_2}\gamma & = &
    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
\end{array}
\]
Since $\tThunk$ serves only as an embedding of negative into positive
types and has no semantic effect,
we interpret thunking and forcing by the identity.
The eliminations for positive types deal now only with values, thus,
need not reference the monad operations.
\[
\begin{array}{lcl}
  \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
    \dent{t}{(\gamma,a_1,a_2)} ) \; \dent v \gamma \\
  \dent{\tcase\,v\,t_1\,t_2}\gamma & = &
    [  \lambda a_1.\ \dent{t_1}{(\gamma,a_1)}
    ,\ \lambda a_2.\ \dent{t_2}{(\gamma,a_2)}
    ]\; \dent v \gamma \\
  \dent{\tabort\,v}\gamma & = & \tmagic\,\dent v \gamma \\
\end{array}
\]
The use of the monad is confined to $\tret$ and $\tbind$.
Note the availability of $\trun^\C : \C \den N \to \den N$
at any negative type $N$ for the interpretation of $\tbind$.
\[
\begin{array}{lcl}
  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
\end{array}
\]

@@

Positive normal forms are values $v : \VNf P \Gamma$
referring only to atomic variables and
whose thunks only
contain negative normal forms.
% Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
% contain negative normal forms.
\begin{gather*}
  \nru{\tvar}{\patom \in \Gamma}{\PValNf {\patom} \Gamma}
  % \nru{\pvar}{P \in \Gamma}{\PValNf P \Gamma}
\qquad
  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValNf{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValNf{P_1}\Gamma \qquad \PValNf{P_2}\Gamma}
      {\PValNfP{P_1 \times P_2}\Gamma}
\\[2ex]
  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
\end{gather*}

Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
starting from a forced $\tThunk$ rather than from variables of negative
types (as those do not exist in CBPV).  However, due to normality
the $\tThunk$ cannot be a $\tthunk$, but only a variable
$\Thunk N \in \Gamma$.
\begin{gather*}
  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeP {N_1 \ntimes N_2} \Gamma}
      {\Ne {N_i} \Gamma}
\\[2ex]
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\end{gather*}
Variables are originally introduced by either $\tabs$ or
the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
$P$.  Variables of composite value type can be broken down by pattern
matching, introducing variables of smaller type.  These positive
eliminations plus $\tbind$ are organized in the
inductively defined strong monad \fbox{$\Cov$}.
In the following rules, parameter $\J$ (for ``judgement'') stands for an
arbitrary presheaf.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tsplit}
      {P_1 \times P_2 \in \Gamma \qquad
       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tcase}
      {P_1 + P_2 \in \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\tabort}
      {\tyempty \in \Gamma}
      {\Cover \J \Gamma}
\end{gather*}
Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
$\Comp P$ is reached.  Then, elimination of neutrals and variables is
possible through the $\Cov$ monad until an answer can be given in form
of a base neutral ($\tNe\,\natom$) or a normal value.
\begin{gather*}
  \nru{\tne}
      {\CoverP {\tNe\,\natom} \Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tret}
      {\CoverP {\tVNf\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\\[2ex]
  \nru{\nunit}{}{\Nf \top \Gamma}
\qquad
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\\[2ex]
  \nru{\tabs}
      {\Nf N {\cext \Gamma P}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
We again can run the cover monad on normal forms, \ie, have
$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
negative semantic values
$\trun : \Cov \den N \todot \den N$.
